Gut-resident microbial species and their genes are associated with cognition and neuroanatomy in children

### 2023-03-28 manuscript/main.tex

As with LMs, RF models for children under 6 months old
performed poorly (mean test-set correlation -0.12, mean RMSE 12.70),
but RFs were consistently able to learn the relationship between taxa
and cognitive function scores in children over 18 months of age (mean
test-set correlation 0.429, mean RMSE 17.29). For both age groups,
species that were important in RF overlapped with those that were
significant when testing the relationship with LMs (Figure
3A-B, Table S1A-B); except for \emph{Eubacterium ramulus}
all taxa that were significant using LMs belonged to a group
responsible for 80\% of the total fitness-weighted relative importance in RF models
(Table 3, Figure 3A).
However, the rank-ordering of LM significance and RF importance
had substantive differences. For example, in children under 6 months,
only \emph{B. wexlerae} was significantly associated with cognitive function in LMs, but
was ranked 20th in importance for RF models.

@@

If there are causal effects of microbial metabolism on cognitive
function, they might be reflected in changes in neuroanatomy. We again
used RF modeling to associate gut taxonomic
profiles with individual brain regions identified in MRI scans.
Some brain regions were more readily
predicted by RF models trained on microbial taxa (Table 4, Table S2), in particular those that were highly correlated with age.
These included the L/R lingual gyrus (mean RF correlations, Left =
0.421, Right = 0.434; relative weighted importance for age, Left = 7.5\%, Right =
8.3\%) and the L/R pericalcarine cortex (mean RF correlations, Left =
0.200, Right = 0.273; relative weighted importance for age, Left = 3.6\%, Right = 6.3\%).
In many cases, however, age was not an important variable in
high-performing models, such as that for the left accumbens area (mean
RF correlations = 0.288; relative age importances = 1.2\%).

@@

While models for many regions were highly concordant across the
left and right hemispheres in terms of both model performance and
microbial feature importance, other regions had substantial
bilateral differences. For example, the left accumbens area
has one of the highest test-set correlations of our
brain region models (R = 0.288), as compared to the right which
had a negative mean test-set correlation (R = -0.041),
indicating that generalizable models were not identified.
Feature importance for models of the left accumbens area were dominated
by three species of \emph{Bacteroides}, \emph{B. vulgatus} (3.3\%
relative weighted importance), \emph{B. ovatus} (3.7\% relative weighted importance), and
\emph{B. uniformis} (2.9\% relative weighted importance).

While RF models for multiple brain regions had many important microbial taxa,
others were dominated by a small number of taxa, many of which
were also identified in LMs and RF models of cognitive function (Figures 2 and 3). 
In general, for all segments, a consistent number of species between 31 and 45 (median = 37, less than 30\% of the input features)
was responsible for half of the fitness-weighted cumulative importance,
regardless of model performance (Figure 4A-B, Table S2).
We observed two major patterns of importance distribution from
species over the brain segments; some species portrayed high contributions
to multiple different segments, while others contributed modestly to
just one or two brain segments. Notable cases of the first pattern
included seven species - \emph{Anaerostipes hadrus}, \emph{B.
vulgatus}, \emph{Fusicatenibacter saccharivorans}, \emph{Ruminococcus
torques}, \emph{Eubacterium rectale}, \emph{Coprococcus comes} and
\emph{B. wexlerae}, which combined, account for approximately 10\% of the cumulative relative importances,
computed after fitness-weighting on the reported brain segments (Figure 4A-B).

@@

Both \emph{A. equolofaciens} and \emph{A. celatus}, two closely-related
equol-producing species, are examples of the second contribution
pattern, where a taxon only contributes significantly to a very small number of segments. 
Each of these species were highly important in predicting the relative volume of
the right anterior cingulate (respectively, 3.0\% and 2.6\% relative weighted
importances). Another example of this pattern is \emph{R. gnavus}, which was the only
species with significant negative association with cognitive function in linear models.
It is heavily associated with the left pars opercularis (2.7\% relative weighted importance). 
Finally, \emph{C. comes} displays an importance
distribution that splits almost evenly among the two previously reported
clusters. Its loading on the prediction of the left posterior cingulate
is the highest for a microbe in RF models (4.0\% relative weighted importance);
only age had higher relative importance in any model. At the same time,
\textit{C. comes} is one of the most important predictors for neighboring areas of the
posterior cingulate such as the pars opercularis (relative weighted importances,
Left = 2.3\%, Right = 2.8\%), along with upper regions like the left
precentral (2.3\% relative weighted importance) and paracentral lobes (relative weighted
importances, Left = 2.6\%, Right = 1.9\%).

@@

To derive biological insight from the models, the covariate variable
importances for all the input features, measured by mean decrease in
impurity (MDI, or GINI importance), was also analyzed. Leveraging the
distribution of results from the extensive repeated cross validation
experiments, rather than electing a representative model or picking the
highest validation-set correlation, a measure of model fitness
(\textbf{Equation 1}) was designed to weight the importances from each
trained forest. The objective was to give more weight to those with
higher benchmarks on the validation sets (or higher generalizability),
while penalizing information from highly overfit models, drawing
inspiration from the approach used on another work employing repeated CV
on Random Forests with high-dimensional, low sample size microbiome
datasets \cite{woodruffInflammationAutoreactivityDefine2022}.
The resulting fitness-weighted importances were used
to generate the values in Figure 3.

>>>>

As with LMs, RF models for children under 6 months old
performed poorly (mean test-set correlation -0.12, mean RMSE 12.70),
but RFs were consistently able to learn the relationship between taxa
and cognitive function scores in children over 18 months of age (mean
test-set correlation 0.429, mean RMSE 17.29). For both age groups,
species that were important in RF overlapped with those that were
significant when testing the relationship with LMs (Figure
3A-B, Table S1A-B); except for \emph{Eubacterium ramulus}
all taxa that were significant using LMs belonged to a group
responsible for 80\% of the total importance in RF models
(Table 3, Figure 3A).
However, the rank-ordering of LM significance and RF importance
had substantive differences. For example, in children under 6 months,
only \emph{B. wexlerae} was significantly associated with cognitive function in LMs, but
was ranked 20th in importance for RF models.

@@

If there are causal effects of microbial metabolism on cognitive
function, they might be reflected in changes in neuroanatomy. We again
used RF modeling to associate gut taxonomic
profiles with individual brain regions identified in MRI scans.
Some brain regions were more readily
predicted by RF models trained on microbial taxa (Table 4, Table S2), in particular those that were highly correlated with age.
These included the L/R lingual gyrus (mean RF correlations, Left =
0.421, Right = 0.434; importance for age, Left = 7.5\%, Right =
8.3\%) and the L/R pericalcarine cortex (mean RF correlations, Left =
0.200, Right = 0.273; importance for age, Left = 3.6\%, Right = 6.3\%).
In many cases, however, age was not an important variable in
high-performing models, such as that for the left accumbens area (mean
RF correlations = 0.288; relative age importances = 1.2\%).

@@

While models for many regions were highly concordant across the
left and right hemispheres in terms of both model performance and
microbial feature importance, other regions had substantial
bilateral differences. For example, the left accumbens area
has one of the highest test-set correlations of our
brain region models (R = 0.288), as compared to the right which
had a negative mean test-set correlation (R = -0.041),
indicating that generalizable models were not identified.
Feature importance for models of the left accumbens area were dominated
by three species of \emph{Bacteroides}, \emph{B. vulgatus} (3.3\% importance),
\emph{B. ovatus} (3.7\% importance), and
\emph{B. uniformis} (2.9\% importance).

While RF models for multiple brain regions had many important microbial taxa,
others were dominated by a small number of taxa, many of which
were also identified in LMs and RF models of cognitive function (Figures 2 and 3). 
In general, for all segments, a consistent number of species between 31 and 45 (median = 37, less than 30\% of the input features)
was responsible for half of the cumulative importance,
regardless of model performance (Figure 4A-B, Table S2).
We observed two major patterns of importance distribution from
species over the brain segments; some species portrayed high contributions
to multiple different segments, while others contributed modestly to
just one or two brain segments. Notable cases of the first pattern
included seven species - \emph{Anaerostipes hadrus}, \emph{B.
vulgatus}, \emph{Fusicatenibacter saccharivorans}, \emph{Ruminococcus
torques}, \emph{Eubacterium rectale}, \emph{Coprococcus comes} and
\emph{B. wexlerae}, which combined, account for approximately 10\% of the cumulative relative importances,
computed after fitness-weighting on the reported brain segments (Figure 4A-B).

@@

Both \emph{A. equolofaciens} and \emph{A. celatus}, two closely-related
equol-producing species, are examples of the second contribution
pattern, where a taxon only contributes significantly to a very small number of segments. 
Each of these species were highly important in predicting the relative volume of
the right anterior cingulate (respectively, 3.0\% and 2.6\% importances).
Another example of this pattern is \emph{R. gnavus}, which was the only
species with significant negative association with cognitive function in linear models.
It is heavily associated with the left pars opercularis (2.7\% importance). 
Finally, \emph{C. comes} has an importance
distribution that splits almost evenly among the two previously reported
clusters. Its loading on the prediction of the left posterior cingulate
is the highest for a microbe in RF models (4.0\% importance);
only age had higher relative importance in any model. At the same time,
\textit{C. comes} is one of the most important predictors for neighboring areas of the
posterior cingulate such as the pars opercularis (importances,
Left = 2.3\%, Right = 2.8\%), along with upper regions like the left
precentral (2.3\% importance) and paracentral lobes (importances, Left = 2.6\%, Right = 1.9\%).

@@

To derive biological insight from the models, the covariate variable
importances for all the input features, measured by mean decrease in
impurity (MDI, or GINI importance), was also analyzed. Leveraging the
distribution of results from the extensive repeated cross validation
experiments, rather than electing a representative model or picking the
highest validation-set correlation, a measure of model fitness
(\textbf{Equation 1}) was designed to weight the importances from each
trained forest, and all "importance" values reported refer to 
the average fitness-weighted importance accross all models.
The objective was to give more weight to those with
higher benchmarks on the validation sets (or higher generalizability),
while penalizing information from highly overfit models, drawing
inspiration from the approach used on another work employing repeated CV
on Random Forests with high-dimensional, low sample size
datasets \cite{woodruffInflammationAutoreactivityDefine2022}.
The resulting fitness-weighted importances were used
to generate the values in Figures 3-4.
